---
title: Power Analysis
subtitle: Simulation-based power analysis for the perceptual validation experiment
---

## Introduction {#sec-power-analysis}

Before conducting the perceptual validation experiment (@sec-experiment), we performed a simulation-based power analysis to determine the minimum sample size required to detect a meaningful effect of the computational model's target values on listeners' perceived flavor ratings.

The analysis targets the linear mixed-effects model:

$$\text{perceived}_{ijk} = \beta_0 + \beta_{\text{target}} \cdot \text{target}_{jk} + u_i + v_j + \varepsilon_{ijk}$$

where $u_i \sim \mathcal{N}(0, \sigma_{\text{subject}}^2)$ is a per-subject random intercept, $v_j \sim \mathcal{N}(0, \sigma_{\text{track}}^2)$ is a per-track random intercept, and $\varepsilon_{ijk} \sim \mathcal{N}(0, \sigma_{\text{residual}}^2)$ is the observation-level residual. The fixed effect $\beta_{\text{target}}$ quantifies the relationship between the model's predicted flavor score and the listener's rating.

### Setup

```{r}
#| label: setup

library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(furrr)

n_cores <- 8
plan(multisession, workers = n_cores)
```

### Simulation Parameters

We set default parameter values based on pilot observations and literature on cross-modal correspondences. Each participant evaluates a random subset of 10 out of 20 tracks, reflecting the incomplete design used in the actual experiment.

```{r}
#| label: parameters

set.seed(123)

N_TRACKS_TOTAL <- 20
N_TRACKS_PER_SUBJECT <- 10
BETA_TARGET <- 0.15
SD_SUBJECT <- 1.2
SD_TRACK <- 1
SD_RESIDUAL <- 1.5
ALPHA <- 0.05

N_SIM <- 100
SAMPLE_SIZES <- seq(20, 120, by = 10)

SD_SUBJECT_GRID <- seq(0.2, 2.0, by = 0.3)
SD_TRACK_GRID <- seq(0.2, 2.0, by = 0.3)
```

### Simulation Functions

The `simulate_dataset()` function generates synthetic data for a given sample size and variance configuration. For each subject, 10 of 20 tracks are randomly selected. Target values are drawn uniformly from [1, 7], and perceived ratings are generated as a linear function of target plus random effects and residual noise.

```{r}
#| label: simulate-function

simulate_dataset <- function(n_subjects, beta_target, sd_subject, sd_track, sd_residual) {
    subjects <- paste0("S", seq_len(n_subjects))
    tracks <- paste0("T", seq_len(N_TRACKS_TOTAL))

    design <- expand_grid(
        subject = subjects,
        track   = tracks
    ) |>
        group_by(subject) |>
        slice_sample(n = N_TRACKS_PER_SUBJECT) |>
        ungroup()

    subject_re <- tibble(
        subject = subjects,
        u_subject = rnorm(n_subjects, 0, sd_subject)
    )

    track_re <- tibble(
        track = tracks,
        u_track = rnorm(N_TRACKS_TOTAL, 0, sd_track)
    )

    design |>
        left_join(subject_re, by = "subject") |>
        left_join(track_re, by = "track") |>
        mutate(
            target = runif(n(), 1, 7),
            perceived = beta_target * target +
                u_subject +
                u_track +
                rnorm(n(), 0, sd_residual)
        )
}
```

The `estimate_power()` function runs `N_SIM` simulations at a given parameter configuration, fits the LMM to each, and returns the proportion of simulations where the target coefficient is significant at $\alpha = 0.05$.

```{r}
#| label: power-function

estimate_power <- function(n_subjects, beta_target, sd_subject, sd_track, sd_residual) {
    p_values <- replicate(N_SIM, {
        data <- simulate_dataset(
            n_subjects, beta_target, sd_subject, sd_track, sd_residual
        )

        model <- suppressMessages(lmer(
            perceived ~ target + (1 | subject) + (1 | track),
            data = data,
            REML = FALSE,
            control = lmerControl(calc.derivs = FALSE)
        ))

        coefs <- summary(model)$coefficients
        if ("target" %in% rownames(coefs)) {
            coefs["target", "Pr(>|t|)"]
        } else {
            NA
        }
    })

    mean(p_values < ALPHA, na.rm = TRUE)
}
```

### Power vs. Sample Size {#sec-power-sample-size}

We evaluate power across sample sizes from 20 to 120 subjects, holding all other parameters at their default values ($\beta_{\text{target}} = 0.15$, $\sigma_{\text{subject}} = 1.2$, $\sigma_{\text{track}} = 1.0$, $\sigma_{\text{residual}} = 1.5$).

```{r}
#| label: fig-power-sample-size
#| fig-cap: "Statistical power as a function of sample size. The dashed horizontal line marks the conventional 80% power threshold."

power_sample_size <- tibble(
    n_subjects = SAMPLE_SIZES
) |>
    mutate(
        power = future_map_dbl(
            n_subjects,
            estimate_power,
            beta_target = BETA_TARGET,
            sd_subject = SD_SUBJECT,
            sd_track = SD_TRACK,
            sd_residual = SD_RESIDUAL,
            .options = furrr_options(seed = TRUE)
        )
    )

ggplot(power_sample_size, aes(x = n_subjects, y = power)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    geom_hline(yintercept = 0.8, linetype = "dashed") +
    labs(
        x = "Number of subjects",
        y = "Statistical power"
    ) +
    theme_minimal()
```

```{r}
#| label: tbl-power-sample-size
#| tbl-cap: "Simulated power by sample size and key operating points for the mixed-effects design."

power_threshold <- 0.8

power_sample_size_table <- power_sample_size |>
    mutate(power = round(power, 3))

power_at_or_above_threshold <- power_sample_size |>
    filter(power >= power_threshold) |>
    arrange(n_subjects)

n_threshold <- if (nrow(power_at_or_above_threshold) > 0) {
    power_at_or_above_threshold$n_subjects[1]
} else {
    NA_integer_
}

power_at_49 <- approx(
    x = power_sample_size$n_subjects,
    y = power_sample_size$power,
    xout = 49,
    rule = 2
)$y

power_operating_points <- tibble(
    metric = c(
        "First N with power >= 0.80",
        "Estimated power at N = 49",
        "Maximum simulated power"
    ),
    value = c(
        ifelse(is.na(n_threshold), "Not reached", as.character(n_threshold)),
        sprintf("%.3f", power_at_49),
        sprintf("%.3f", max(power_sample_size$power))
    )
)

power_sample_size_table
power_operating_points
```

#### Power vs. Random-Effect Variance {#sec-power-variance}

We fix the sample size at $N = 50$ and vary both subject-level and track-level standard deviations across a grid from 0.2 to 2.0. This characterizes how sensitive the design is to individual differences and track-level variability.

```{r}
#| label: fig-power-variance
#| fig-cap: "Statistical power as a function of subject and track random-effect standard deviations, at fixed N = 50."

power_variance <- expand_grid(
    sd_subject = SD_SUBJECT_GRID,
    sd_track   = SD_TRACK_GRID
) |>
    mutate(
        power = future_pmap_dbl(
            list(sd_subject, sd_track),
            ~ estimate_power(
                n_subjects = 50,
                beta_target = BETA_TARGET,
                sd_subject = ..1,
                sd_track = ..2,
                sd_residual = SD_RESIDUAL
            ),
            .options = furrr_options(seed = TRUE)
        )
    )

ggplot(power_variance, aes(x = sd_subject, y = sd_track, fill = power)) +
    geom_tile() +
    scale_fill_viridis_c(limits = c(0, 1)) +
    labs(
        x = "Subject SD",
        y = "Track SD",
        fill = "Power"
    ) +
    theme_minimal()
```

```{r}
#| label: tbl-power-variance-summary
#| tbl-cap: "Summary of the variance-sensitivity grid at fixed N = 50."

power_variance_summary <- power_variance |>
    summarise(
        min_power = min(power),
        median_power = median(power),
        max_power = max(power),
        pct_ge_80 = mean(power >= 0.8) * 100
    ) |>
    mutate(across(everything(), ~ round(.x, 3)))

power_variance_extremes <- bind_rows(
    power_variance |>
        arrange(power) |>
        slice_head(n = 3) |>
        mutate(case = "Lowest power"),
    power_variance |>
        arrange(desc(power)) |>
        slice_head(n = 3) |>
        mutate(case = "Highest power")
) |>
    mutate(power = round(power, 3))

power_variance_summary
power_variance_extremes
```

```{r}
#| label: cleanup
plan(sequential)
```

## Interpretation

The power curve (@fig-power-sample-size) quantifies how power grows with sample size under the default simulation assumptions. In this setup, the first simulated sample size that reaches the 0.80 threshold is $N=`r n_threshold`$ (Table @tbl-power-sample-size), and the estimated power at the final experimental sample size ($N=49$) is `r sprintf('%.3f', power_at_49)`.

The variance sensitivity analysis (@fig-power-variance) shows that power remains strongly dependent on random-effect variability. Across the full grid at fixed $N=50$, power ranges from `r sprintf('%.3f', power_variance_summary$min_power)` to `r sprintf('%.3f', power_variance_summary$max_power)`, with median `r sprintf('%.3f', power_variance_summary$median_power)` and `r sprintf('%.1f', power_variance_summary$pct_ge_80)`\% of conditions at or above 0.80 (Table @tbl-power-variance-summary). Together, these results support the sample-size choice used in the perceptual validation study reported in @sec-experiment.
